---
title: "Chicago Hospital WGS metadata"
output: html_document
date: "2023-06-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Metadata from SRA project PRJEB13117

The full metadata table is very tedious to look through. I want to pick out the columns that might be informative in reanalyzing the data to save as a smaller table. Along the way, I want to get an idea of what I have to work with.

```{r}
suppressPackageStartupMessages(library(dplyr))
library(tidyr)
library(ggplot2)
```

```{r}
data = read.csv("PRJEB13117_SraRunTable.csv")
row.names(data) = data$Run
```

Save any output to this folder:
```{r}
outDir = "PRJEB13117_output"
suppressWarnings(dir.create(outDir))
```

## General sorting of columns

Weed through the un-interesting things
```{r}

# columnes with the same value for all rows
allRowCols = c()
allRowVals = c()

# columns with unique values in each row
pot.identifier =c()

# known to be boring adminstriative info
boring = c("create_date", "DATASTORE.filetype", "DATASTORE.provider", "DATASTORE.region")

for (col in names(data)){
  vals = unique(data[,col])
  if(col %in% boring) a=1
  else if (length(vals)==1){
    allRowCols = union(allRowCols, col)
    allRowVals = c(allRowVals, paste0("all rows: ", col, " = ", paste(vals, collapse = ", ")))
  } 
  else if (length(vals) == nrow(data) ){
    pot.identifier = union(pot.identifier, col)
    message("potentially an id column: ", col, " (n=", length(vals), ")")
    # print(vals[1:6])
  }
  else if (length(vals) > 6 ) {
    message(col, " (n=", length(vals), ")")
    print(vals[1:6])
  }else {
    message(col, " (n=", length(vals), ")")
    print(vals)
  }
}
```

It looks like there are several id columns; though not "Bytes" or "Bases".

It looks like our only informative column with different values between samples is Collection_Date.

Several columns have valures that are the same for all rows, and some of them are invormative about the dataset as a whole, but I don't need to keep a whole column of repeated data.


## Whats interesting

```{r}
data$date = as.Date(data$Collection_Date)
data$thDay = data$date - min(data$date)
datesdf = data.frame(table(data$date))
names(datesdf) = c("Collection_Date", "numberSamples")
write.table(datesdf, file=file.path(outDir, "samplesPerDate.txt"),
            quote=F, row.names = F, sep="\t")
datesdf
```

Previewing the data that's the same for alls rows, bits that I think are noteworthy are these:
```{r}
usefulBits = c(
  "Center.Name = NASEER",
  "assay.Type = WGS",
  "AvgSpotLen = 202",
  "Organism = metagenome"         ,                    
  "Instrument = Illumina HiSeq 1500",
  "SRA.Study = ERP014653",
  "ventilation_type = mechanical ventilation",
  "filter_type = electrostatic air treatment",
  "LibraryLayout = PAIRED",                                     
  "LibrarySelection = RANDOM PCR",                            
  "LibrarySource = METAGENOMIC"
)
```

## Write reduced version of the whole table.

Of the several id columns, we can just keep Run, that id can be used to link back if needed; none of the others look immediately usefull.

```{r}
file = file.path(outDir, "PRJEB13117_SraRunTable_reduced.txt")
keep_cols = c("Run", "Bytes", "Collection_Date", "thDay")
writeLines(file, text=paste("#", usefulBits))
write.table(data[,keep_cols], file=file,
            quote=F, row.names = F, append = T, sep="\t")
```


## General review of columns

Is anything else interesting?
```{r}
for (col in names(data)){
  vals = unique(data[,col])
  if (length(vals)==1){
    message("all rows: ", col, " = ", vals)
  } 
  else if (length(vals) > 6 ) {
    message(col, " (n=", length(vals), ")")
    print(vals[1:6])
  }else {
    message(col, " (n=", length(vals), ")")
    print(vals)
  }
}
```



```{r}
sessionInfo()
```

