---
title: "Chicago Hospital Metadata"
output: html_document
date: "2023-06-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Metadata from SRA project PRJEB14474

The full metadata table is very tedious to look through. I want to pick out the columns that might be informative in reanalyzing the data to save as a smaller table. Along the way, I want to get an idea of what I have to work with.

```{r}
suppressPackageStartupMessages(library(dplyr))
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggrepel)
```

```{r}
data = read.delim2("../../data/NOT_IN_GIT/meta/PRJEB14474_SraRunTable.txt")
row.names(data) = data$ID
```

Save any output to this folder:
```{r}
outDir = "../output"
suppressWarnings(dir.create(outDir))
```


## Some modified data

### Date

Convert the Date values into sortable numeric values.
```{r}
ds = strsplit(data$Date, split="/")
ds2 = do.call(rbind, ds) %>% data.frame()
row.names(ds2) = data$ID
names(ds2) = c("month", "day", "year")
ds2$monthC = as.character(ds2$month)
need0 = as.numeric(ds2$month) < 10
ds2$monthC[need0] = paste0("0", ds2$month[need0])
ds2$datestring = paste0("20", ds2$year, "-", ds2$monthC, "-", ds2$day)
ds2$date = as.Date(ds2$datestring)
ds2$thDay = as.numeric( ds2$date - min(ds2$date) )
```

Dates are represented as number of days after `r min(ds2$date)` in the column `study_day`

```{r}
head(ds2)
```


### host_subject_id

This id column seems to have 3 parts separated by ".".  

```{r}
head(data$host_subject_id)
```

First looks like location, second looks like sampel type, and third is an integer.  Not sure what that means.  Lets grab it and check it out.

```{r}
data$hsi3 = sapply(strsplit(data$host_subject_id, split=".", fixed=T),
             "[", 3)
data$hsi3 = as.numeric(data$hsi3)
summary(data$hsi3)
```

Lets see if it is correlated with any other data.

```{r}
ds3 = merge(ds2, data["hsi3"], by=0)
row.names(ds3) = ds3$Row.names
ds3 = ds3[,grep(x=names(ds3), "Row.names", value=T, invert = T)]
plot(ds3$hsi3 , ds3$thDay, las=1)
```
```{r}
table(ds3$thDay - ds3$hsi3 == 81)
```

Oh! I guess this number refers to integer day.

And maybe 0 is the day the hospital opened?  The paper says "Sampling began 2 months before the hos- pital opening on 23 February 2013 and continued for a year thereafter."

```{r}
is0 = which(ds3$hsi3 == 0 )
ds3$date[is0]
```

Close enough! Lets call this "study_day".
```{r}
ds3$study_day = ds3$hsi3
```

## Reduced data frame

Select limited columns.
```{r}
df = merge(data, ds3, by=0)
df2 = df[,c("ID", "Sample.Name", "Sample_name", "host_subject_id", "Date", "study_day", "SAMPLE_TYPE", "Location", "run_prefix..exp.", "env_material", "surcat", "Bases", "AvgSpotLen")]
```

Give a value to the blank run prefix.
```{r}
blankVals = df2$run_prefix..exp. == ""
df2$run_prefix..exp.[blankVals] = "otherRun"
```

```{r}
write.table(file = file.path(outDir, "PRJEB14474_SraRunTable_reduced.txt"), 
            df2, quote=F, row.names = F, sep="\t")
```


## Visually summarize

### Whats interesting


Location vs SAMPLE_TYPE
```{r}
df7 = df2 %>% 
  mutate(SAMPLE_TYPE = paste0("(", surcat, ") ", SAMPLE_TYPE)) %>%
  count(SAMPLE_TYPE, Location)

hm = ggplot(df7, aes(x=Location, y=SAMPLE_TYPE, color=n, fill=n)) + 
  geom_tile() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  ggtitle("PRJEB14474 samples", subtitle = paste("N =", sum(df7$n)))
hm

ggsave(file.path(outDir, "Locatoin-sampletype_heat.png"), hm)
write.table(df7 %>% 
              mutate_all(function(x) replace(x, x=='', NA)) %>%
              pivot_wider(names_from = Location, values_from = "n" ),
            file = file.path(outDir, "Location-sampletype_table.txt"),
            quote=F, sep="\t", row.names = F)
```
I guess the 'blank control' samples are not assigned a location.

Repeat, but only consider unque samples.
```{r}
df8 = df2 %>% 
  filter(!duplicated(Sample.Name)) %>%
  mutate(SAMPLE_TYPE = paste0("(", surcat, ") ", SAMPLE_TYPE)) %>%
  count(SAMPLE_TYPE, Location)

hm = ggplot(df8, aes(x=Location, y=SAMPLE_TYPE, color=n, fill=n)) + 
  geom_tile() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  ggtitle("PRJEB14474 samples (unqiue samples)", subtitle = paste("N =", sum(df8$n)))
hm
```

That looks pretty much the same. Where are the technical replicates?

```{r}
df8 = df2 %>% 
  filter(duplicated(Sample.Name)) %>%
  mutate(SAMPLE_TYPE = paste0("(", surcat, ") ", SAMPLE_TYPE)) %>%
  count(SAMPLE_TYPE, Location)

hm = ggplot(df8, aes(x=Location, y=SAMPLE_TYPE, color=n, fill=n)) + 
  geom_tile() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  ggtitle("PRJEB14474 samples (technical replicates)", subtitle = paste("N =", sum(df8$n)))
hm
```



Sampling over time.

Visually, how many of each type of sample were collected each day?
```{r}
df3 = df2 %>% count(study_day, SAMPLE_TYPE)
# ggplot(df3) +
#   geom_tile(aes(x=study_day, y=SAMPLE_TYPE, color=n, fill=n)) +
#   scale_color_gradient(low="blue", high="red") +
#   scale_fill_gradient(low="blue", high="red")
```

That's so fine its hard to read. 

What if we lump together samples collected in a 10-day period?
```{r}
df4 = df2 %>% 
  mutate(SAMPLE_TYPE = paste0("(", surcat, ") ", SAMPLE_TYPE)) %>%
  mutate(ishDay=round(study_day, -1)) %>% 
  count(ishDay, SAMPLE_TYPE)

byday = ggplot(df4) +
  geom_tile(aes(x=ishDay, y=SAMPLE_TYPE, color=n, fill=n)) +
  scale_color_gradient(low="blue", high="red") +
  scale_fill_gradient(low="blue", high="red") +
  xlab("Day (rounded to 10-day period)") +
  geom_vline(x = 0,
             xintercept = 0,
             linetype = "solid") +
  annotate(geom = "text",
             label = " opening",
             x = 0, y=-1, 
             vjust = 1, hjust=0)
# +
  # coord_cartesian(ylim = c(0, 23), clip = "off")
byday

ggsave(file.path(outDir, "time-sampletype_heat.png"), byday)
```



For each sample type, how are samples distributed over time?
```{r}
p2 = ggplot(df3, aes(x=study_day, y=n)) +
  geom_point() +
  facet_wrap(facets="SAMPLE_TYPE") +
  geom_point(aes(color=as.character(n))) + scale_color_hue()
p2
```

Location sampling over time

```{r}
df9 = df2 %>% 
  mutate(ishDay=round(study_day, -1)) %>% 
  count(ishDay, Location)

byday = ggplot(df9) +
  geom_tile(aes(x=ishDay, y=Location, color=n, fill=n)) +
  scale_color_gradient(low="blue", high="orange") +
  scale_fill_gradient(low="blue", high="orange") +
  xlab("Day (rounded to 10-day period)") +
  geom_vline(x = 0,
             xintercept = 0,
             linetype = "solid") +
  annotate(geom = "text",
             label = " opening",
             x = 0, y=-1, 
             vjust = 1, hjust=0)
byday

ggsave(file.path(outDir, "time-location_heat.png"), byday)
```
### Bases per sample type

```{r}
data %>% 
  count(AvgSpotLen, run_prefix..exp.) %>%
  mutate_at(c("AvgSpotLen", "run_prefix..exp."), function(x) replace(x, x=='', "unknown")) %>%
  pivot_wider(names_from = run_prefix..exp., values_from= n)

p1 = ggplot(data, aes(x=SAMPLE_TYPE, y=Bases, color=surcat)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) 

p1 + geom_boxplot()
```
Of the subset that I used for testing, which samples have the most bases?
```{r}
subset = df2 %>% filter(df$study_day==4)
subsetPlot = ggplot(subset, aes(x=SAMPLE_TYPE, y=Bases, color=surcat)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) 
# subsetPlot + geom_text(aes(label=ID, vjust=-.5))
# subsetPlot + geom_label(aes(label=ID, vjust=-.5))
subsetPlot + geom_text_repel(aes(label = ID), show.legend = FALSE)
# three cheers for the ggrepel package! otherwise the labels for ERR1460012 and ERR1459325 would cover each other!
```

Well, fiddles. I gave the intern a subset where the blanks are some of the largest samples and some of the actual samples are essentially empty.


## Which columns are redundant

### SAMPLE_TYPE = surface

Is there any difference between the "SAMPLE_TYPE" column and the "surface" column?
```{r}
ns = which(data$surface != data$SAMPLE_TYPE)
cotype = paste(data$surface[ns], data$SAMPLE_TYPE[ns])
table(cotype)
```
No. SAMPLE_TYPE and surface are the same. One uses the term BLANK the other says "blank control", otherwise identical.

What are the values of the "SAMPLE_TYPE" feature? How often does each value appear?
```{r}
table(data$SAMPLE_TYPE)
```

There are 22 values. "glove from box" might be considered additional blanks.

### Location = loccat

Is there any difference between "Location" and "loccat"?
```{r}
table(data$Location == data$loccat)
```
No. Those are synonymous features.

What are the values for Location? How often do they appear?
```{r}
table(data$Location)
```
### run_prefix and run_date

Is there a relationship between "run_prefix..exp." and "target_gene..exp." ?
```{r}
data %>% 
  count(target_gene..exp., run_prefix..exp.) %>%
  mutate_at(c("target_gene..exp.", "run_prefix..exp."), function(x) replace(x, x=='', "unknown")) %>%
  pivot_wider(names_from = run_prefix..exp., values_from= n)
```
This is probably just a data-entry artifact.  Only the run_prefix is meaningful--if it really different sequencing.
Notice that "target_subfragment..exp." is always V4.

Is run_date..exp. interesting?
```{r}
dats = split(data, f=data$run_date..exp.)
# lapply(dats, function(df) unique(df$SAMPLE_TYPE))
# lapply(dats, function(df) unique(df$Location))
lapply(dats, function(df) unique(df$run_prefix..exp.))
# lapply(dats, function(df) unique(df$Date) %>% sort())
```
I had assumed run_date would be the date the sample was run on the sequencer.  I wonder if a subset of samples was a different data entry batch so they have "16S rRNA" instead of "16S", and they don't specify a run_prefix and the date was entered incorrectly (could be the date was set incorrectly in the sequencer?).

Does run_prefix have any relationship with Location? 
```{r}
data %>% 
  count(Location, run_prefix..exp.) %>%
  mutate_at(c("Location", "run_prefix..exp."), function(x) replace(x, x=='', "unknown")) %>%
  pivot_wider(names_from = run_prefix..exp., values_from= n)
```

Does run_prefix have any relationship with SAMPLE_TYPE?
```{r}
data %>% 
  count(SAMPLE_TYPE, run_prefix..exp.) %>%
  mutate_at(c("SAMPLE_TYPE", "run_prefix..exp."), function(x) replace(x, x=='', "unknown")) %>%
  pivot_wider(names_from = run_prefix..exp., values_from= n)
```

Does run_prefix have any relationship with AvgSpotLen?
```{r}
data %>% 
  count(AvgSpotLen, run_prefix..exp.) %>%
  mutate_at(c("AvgSpotLen", "run_prefix..exp."), function(x) replace(x, x=='', "unknown")) %>%
  pivot_wider(names_from = run_prefix..exp., values_from= n)
```
The samples with prefix "MiSeq" have avg spot lengths of 150-151. All of the ones that start with "HiSeq" have ave spot length <150.  I'm not sure why there is a relationship between different HiSeq prefix sets and different spot lengths.

### Date = COLLECTION_TIMESTAMP

Is there any difference between "Date" and "COLLECTOIN_TIMESTAMP"
```{r}
table(data$Date == gsub(" 12:00", "", data$COLLECTION_TIMESTAMP))
```
COLLECTOIN_TIMESTAMP includes " 12:00" at the end of the Date value.



### run_prefix vs Date

Does run_prefix have any relationship with Date?
```{r}
df5 = df2 %>% 
  mutate_at(c("study_day"), function(x) round(x, -1)) %>%
  count(study_day, run_prefix..exp.) %>%
  mutate_at(c("study_day", "run_prefix..exp."), function(x) replace(x, x=='', "unknown")) 

ggplot(df5, aes(x=as.numeric(study_day), y=run_prefix..exp., color=n, fill=n)) + 
  geom_tile()
```

```{r}
# df5 %>% pivot_wider(names_from = run_prefix..exp., values_from= n)
```

### env_biome

Is env_biome interesting?
```{r}
table(data$env_biome)
```

Nope. not interesting.

### env_feature

Is env_feature interesting?
```{r}
feats = split(data, f=data$env_feature)
lapply(feats, function(df) unique(df$SAMPLE_TYPE))
```
No, env_feature is a grouping or category of SAMPLE_TYPE; SAMPLE_TYPE is a sub-type of env_feature. not interesting.

### env_material

Is env_material interesting?
```{r}
mats = split(data, f=data$env_material)
lapply(mats, function(df) unique(df$SAMPLE_TYPE))
```
Not really, env_material is a grouping or category of SAMPLE_TYPE; SAMPLE_TYPE is a sub-type of env_material not interesting. But, a helpful category, so lets keep it.

### surcat

Is surcat interesting?
```{r}
surcats = split(data, f=data$surcat)
lapply(surcats, function(df) unique(df$SAMPLE_TYPE))
```
Yes! surcat is interesting because it distinguishes staff skin from patient skin for "Hand" and "Nose" samples. (that was implied by location...but not explicit)

Does that correspond with Location?
```{r}
lapply(surcats[c("Patient_Skin", "Staff_Skin")], function(df) unique(df$Location))
```
yes! 

## General review of columns

Is anything else interesting?
```{r}
for (col in names(data)){
  vals = unique(data[,col])
  if (length(vals)==1){
    message("all rows: ", col, " = ", vals)
  } 
  else if (length(vals) > 6 ) {
    message(col, " (n=", length(vals), ")")
    print(vals[1:6])
  }else {
    message(col, " (n=", length(vals), ")")
    print(vals)
  }
}
```

I think we grabbed all the info that we might care about. :)


```{r}
sessionInfo()
```

