---
title: "Chicago Hospital Metadata"
output: html_document
date: "2023-06-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Metadata from SRA project PRJEB14474

The full metadata table is very tedious to look through. I want to pick out the columns that might be informative in reanalyzing the data to save as a smaller table. Along the way, I want to get an idea of what I have to work with.

```{r}
suppressPackageStartupMessages(library(dplyr))
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggrepel)
```

```{r}
data = read.delim2("../../data/NOT_IN_GIT/meta/PRJEB14474_SraRunTable.txt")
row.names(data) = data$ID
```

Save any output to this folder:
```{r}
outDir = "../output"
suppressWarnings(dir.create(outDir))
```


## Some modified data

### Date

Convert the Date values into sortable numeric values.
```{r}
ds = strsplit(data$Date, split="/")
ds2 = do.call(rbind, ds) %>% data.frame()
row.names(ds2) = data$ID
names(ds2) = c("month", "day", "year")
ds2$monthC = as.character(ds2$month)
need0 = as.numeric(ds2$month) < 10
ds2$monthC[need0] = paste0("0", ds2$month[need0])
ds2$datestring = paste0("20", ds2$year, "-", ds2$monthC, "-", ds2$day)
ds2$date = as.Date(ds2$datestring)
ds2$thDay = as.numeric( ds2$date - min(ds2$date) )
```

Dates are represented as number of days after `r min(ds2$date)` in the column `study_day`

```{r}
head(ds2)
```


### host_subject_id

This id column seems to have 3 parts separated by ".".  

```{r}
head(data$host_subject_id)
```

First looks like location, second looks like sampel type, and third is an integer.  Not sure what that means.  Lets grab it and check it out.

```{r}
data$hsi3 = sapply(strsplit(data$host_subject_id, split=".", fixed=T),
             "[", 3)
data$hsi3 = as.numeric(data$hsi3)
summary(data$hsi3)
```

Lets see if it is correlated with any other data.

```{r}
ds3 = merge(ds2, data["hsi3"], by=0)
row.names(ds3) = ds3$Row.names
ds3 = ds3[,grep(x=names(ds3), "Row.names", value=T, invert = T)]
plot(ds3$hsi3 , ds3$thDay, las=1)
```
```{r}
table(ds3$thDay - ds3$hsi3 == 81)
```

Oh! I guess this number refers to integer day.

And maybe 0 is the day the hospital opened?  The paper says "Sampling began 2 months before the hos- pital opening on 23 February 2013 and continued for a year thereafter."

```{r}
is0 = which(ds3$hsi3 == 0 )
ds3$date[is0]
```

Close enough! Lets call this "study_day".
```{r}
ds3$study_day = ds3$hsi3
```

## Reduced data frame

Select limited columns.
```{r}
df = merge(data, ds3, by=0)
df2 = df[,c("ID", "Sample.Name", "Sample_name", "host_subject_id", "Date", "study_day", "SAMPLE_TYPE", "Location", "run_prefix..exp.", "env_material", "surcat", "Bases", "AvgSpotLen")]
```

Give a value to the blank run prefix.
```{r}
blankVals = df2$run_prefix..exp. == ""
df2$run_prefix..exp.[blankVals] = "otherRun"
```

```{r}
write.table(file = file.path(outDir, "PRJEB14474_SraRunTable_reduced.txt"), 
            df2, quote=F, row.names = F, sep="\t")
```


## Visually summarize

### Whats interesting


Location vs SAMPLE_TYPE
```{r}
df7 = df2 %>% 
  mutate(SAMPLE_TYPE = paste0("(", surcat, ") ", SAMPLE_TYPE)) %>%
  count(SAMPLE_TYPE, Location)

hm = ggplot(df7, aes(x=Location, y=SAMPLE_TYPE, color=n, fill=n)) + 
  geom_tile() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  ggtitle("PRJEB14474 samples", subtitle = paste("N =", sum(df7$n)))
hm

ggsave(file.path(outDir, "Locatoin-sampletype_heat.png"), hm)
write.table(df7 %>% 
              mutate_all(function(x) replace(x, x=='', NA)) %>%
              pivot_wider(names_from = Location, values_from = "n" ),
            file = file.path(outDir, "Location-sampletype_table.txt"),
            quote=F, sep="\t", row.names = F)
```
I guess the 'blank control' samples are not assigned a location.

Repeat, but only consider unque samples.
```{r}
df8 = df2 %>% 
  filter(!duplicated(Sample.Name)) %>%
  mutate(SAMPLE_TYPE = paste0("(", surcat, ") ", SAMPLE_TYPE)) %>%
  count(SAMPLE_TYPE, Location)

hm = ggplot(df8, aes(x=Location, y=SAMPLE_TYPE, color=n, fill=n)) + 
  geom_tile() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  ggtitle("PRJEB14474 samples (unqiue samples)", subtitle = paste("N =", sum(df8$n)))
hm
```

That looks pretty much the same. Where are the technical replicates?

```{r}
df8 = df2 %>% 
  filter(duplicated(Sample.Name)) %>%
  mutate(SAMPLE_TYPE = paste0("(", surcat, ") ", SAMPLE_TYPE)) %>%
  count(SAMPLE_TYPE, Location)

hm = ggplot(df8, aes(x=Location, y=SAMPLE_TYPE, color=n, fill=n)) + 
  geom_tile() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  ggtitle("PRJEB14474 samples (technical replicates)", subtitle = paste("N =", sum(df8$n)))
hm
```



Sampling over time.

Visually, how many of each type of sample were collected each day?
```{r}
df3 = df2 %>% count(study_day, SAMPLE_TYPE)
# ggplot(df3) +
#   geom_tile(aes(x=study_day, y=SAMPLE_TYPE, color=n, fill=n)) +
#   scale_color_gradient(low="blue", high="red") +
#   scale_fill_gradient(low="blue", high="red")
```

That's so fine its hard to read. 

What if we lump together samples collected in a 10-day period?
```{r}
df4 = df2 %>% 
  mutate(SAMPLE_TYPE = paste0("(", surcat, ") ", SAMPLE_TYPE)) %>%
  mutate(ishDay=round(study_day, -1)) %>% 
  count(ishDay, SAMPLE_TYPE)

byday = ggplot(df4) +
  geom_tile(aes(x=ishDay, y=SAMPLE_TYPE, color=n, fill=n)) +
  scale_color_gradient(low="blue", high="red") +
  scale_fill_gradient(low="blue", high="red") +
  xlab("Day (rounded to 10-day period)") +
  geom_vline(xintercept = 0,
             linetype = "solid") +
  annotate(geom = "text",
             label = " opening",
             x = 0, y=-1, 
             vjust = 1, hjust=0)
# +
  # coord_cartesian(ylim = c(0, 23), clip = "off")
byday

ggsave(file.path(outDir, "time-sampletype_heat.png"), byday)
```



For each sample type, how are samples distributed over time?
```{r}
p2 = ggplot(df3, aes(x=study_day, y=n)) +
  geom_point() +
  facet_wrap(facets="SAMPLE_TYPE") +
  geom_point(aes(color=as.character(n))) + scale_color_hue()
p2
```

Location sampling over time

```{r}
df9 = df2 %>% 
  mutate(ishDay=round(study_day, -1)) %>% 
  count(ishDay, Location)

byday = ggplot(df9) +
  geom_tile(aes(x=ishDay, y=Location, color=n, fill=n)) +
  scale_color_gradient(low="blue", high="orange") +
  scale_fill_gradient(low="blue", high="orange") +
  xlab("Day (rounded to 10-day period)") +
  geom_vline(x = 0,
             xintercept = 0,
             linetype = "solid") +
  annotate(geom = "text",
             label = " opening",
             x = 0, y=-1, 
             vjust = 1, hjust=0)
byday

ggsave(file.path(outDir, "time-location_heat.png"), byday)
```

Specifically for the patient surface samples, for each type, what location / time distribution do I have? Does it look like it might be enough into to infer when the patient changes?
```{r}
colorScale = c("blue", "orange", "darkgreen", "pink", "hotpink", "darkred")

df9.2 = df2 %>% 
  filter(surcat == "Patient_Skin") %>% 
  #filter(SAMPLE_TYPE == "Hand") %>% 
  count(study_day, Location, SAMPLE_TYPE) %>%
  mutate(N=as.factor(n))

byday2 = ggplot(df9.2) +
  geom_tile(aes(x=study_day, y=Location, fill=N)) +
  scale_fill_discrete(type=colorScale) +
  xlab("study day") +
  # theme_get() +
  facet_wrap("SAMPLE_TYPE", ncol = 1)

ggsave(file.path(outDir, "patient-skin_time.v.location_sample-count.png"), byday2)

byday2
```
A lot of those, look like the different sample types were taken on the same day. Plot them all together to see if they do indeed line up.
```{r}
df9.1 = df2 %>%
  filter(surcat == "Patient_Skin") %>%
  count(study_day, Location) %>%
  mutate(N=as.factor(n))

byday1 = ggplot(df9.1) +
  geom_tile(aes(x=study_day, y=Location, fill=N)) +
  scale_fill_discrete(type=colorScale) +
  xlab("study day") +
  theme_minimal()
byday1
```

Yup, as expected, the gaps from the individual samples types are still the gaps when they are all together.

Save one iteration that is just the plot for the hand data.
```{r}
df9.hand = df2 %>% 
  filter(SAMPLE_TYPE == "Hand") %>%
  count(study_day, Location, SAMPLE_TYPE) %>%
  mutate(N=as.factor(n))

byday4 = ggplot(df9.2) +
  geom_tile(aes(x=study_day, y=Location, fill=N)) +
  scale_fill_discrete(type=colorScale) +
  xlab("study day") +
  ggtitle("Patient Hand Sample Counts")

ggsave(file.path(outDir, "patient-hand_time.v.location_sample-count.png"), byday4, height = 2.2)

byday4
```


### Bases per sample type

```{r}
data %>% 
  count(AvgSpotLen, run_prefix..exp.) %>%
  mutate_at(c("AvgSpotLen", "run_prefix..exp."), function(x) replace(x, x=='', "unknown")) %>%
  pivot_wider(names_from = run_prefix..exp., values_from= n)

p1 = ggplot(data, aes(x=SAMPLE_TYPE, y=Bases, color=surcat)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) 

p1 + geom_boxplot()
```
Of the subset that I used for testing, which samples have the most bases?
```{r}
subset = df2 %>% filter(df$study_day==4)
subsetPlot = ggplot(subset, aes(x=SAMPLE_TYPE, y=Bases, color=surcat)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) 
# subsetPlot + geom_text(aes(label=ID, vjust=-.5))
# subsetPlot + geom_label(aes(label=ID, vjust=-.5))
subsetPlot + geom_text_repel(aes(label = ID), show.legend = FALSE)
# three cheers for the ggrepel package! otherwise the labels for ERR1460012 and ERR1459325 would cover each other!
```

Well, fiddles. I gave the intern a subset where the blanks are some of the largest samples and some of the actual samples are essentially empty.

## Compare Lax et al. 2017 Table S1

From the supplemental data file:
 > Table S1. Summary of the 6523 samples included in the study, grouped by surface type.
 
```{r}
preCount = df2 %>% 
  filter(study_day < 0 ) %>%
  count(surcat, SAMPLE_TYPE) %>%
  mutate(sra.pre = n) %>%
  select(-n)
postCount = df2 %>% 
  filter(study_day >= 0 ) %>%
  count(surcat, SAMPLE_TYPE) %>%
  mutate(sra.post = n) %>%
  select(-n)
totCount = df2 %>% 
  count(surcat, SAMPLE_TYPE) %>%
  mutate(sra.total = n) %>%
  select(-n)

m2 = merge(preCount, postCount, all = T)
m2 = merge(m2, totCount, all = T)
m2[m2$SAMPLE_TYPE=="blank control","surcat"] = "Blank"

s1 = read.delim2("../../data/inGit/Lax-et-al.2017_TableS1.txt")
m3 = merge(m2, s1, all = T)
m3[is.na(m3)] = 0

m3$missing.pre = m3$Pre_Opening - m3$sra.pre
m3$missing.post = m3$Post_Opening - m3$sra.post
m3$missing.tot = m3$Surface_Total - m3$sra.total

write.table(m3, file=file.path(outDir, "SampleCountByType_sra-vs-tableS1.txt"), 
                               sep="\t", row.names=F, quote=F)

m3
```

It can be hard to match up exactly what is missing.  We expected to find 6,523 samples, and only have 3,079.  Keep in mind that some of the sra table rows are technical replicates.

Based on this matchup, we see `r sum(s1$Surface_Total)` described in table S1, which comes in as `r sum(m3$Surface_Total)` in the joined table. 
We see `r nrow(meta)` described in the sra metadata, which comes in as `r sum(m3$sra.total)` in the joined table.  We are missing `r sum(m3$missing.tot)` samples.

It looks like we have more "Inguinal Fold" samples than we expect (?). This might be explained by technical replicates not actually counting as unique samples.
```{r}
length(unique(df2$Sample.Name[df2$SAMPLE_TYPE=="Inguinal Fold"]))
```

Table S1 says there are 28 Inuinal Fold samples. We have 36 runs of this sample type, but only 22 of those are actually unique samples.

## Which columns are redundant

### SAMPLE_TYPE = surface

Is there any difference between the "SAMPLE_TYPE" column and the "surface" column?
```{r}
ns = which(data$surface != data$SAMPLE_TYPE)
cotype = paste(data$surface[ns], data$SAMPLE_TYPE[ns])
table(cotype)
```
No. SAMPLE_TYPE and surface are the same. One uses the term BLANK the other says "blank control", otherwise identical.

What are the values of the "SAMPLE_TYPE" feature? How often does each value appear?
```{r}
table(data$SAMPLE_TYPE)
```

There are 22 values. "glove from box" might be considered additional blanks.

### Location = loccat

Is there any difference between "Location" and "loccat"?
```{r}
table(data$Location == data$loccat)
```
No. Those are synonymous features.

What are the values for Location? How often do they appear?
```{r}
table(data$Location)
```
### run_prefix and run_date

Is there a relationship between "run_prefix..exp." and "target_gene..exp." ?
```{r}
data %>% 
  count(target_gene..exp., run_prefix..exp.) %>%
  mutate_at(c("target_gene..exp.", "run_prefix..exp."), function(x) replace(x, x=='', "unknown")) %>%
  pivot_wider(names_from = run_prefix..exp., values_from= n)
```
This is probably just a data-entry artifact.  Only the run_prefix is meaningful--if it really different sequencing.
Notice that "target_subfragment..exp." is always V4.

Is run_date..exp. interesting?
```{r}
dats = split(data, f=data$run_date..exp.)
# lapply(dats, function(df) unique(df$SAMPLE_TYPE))
# lapply(dats, function(df) unique(df$Location))
lapply(dats, function(df) unique(df$run_prefix..exp.))
# lapply(dats, function(df) unique(df$Date) %>% sort())
```
I had assumed run_date would be the date the sample was run on the sequencer.  I wonder if a subset of samples was a different data entry batch so they have "16S rRNA" instead of "16S", and they don't specify a run_prefix and the date was entered incorrectly (could be the date was set incorrectly in the sequencer?).

Does run_prefix have any relationship with Location? 
```{r}
data %>% 
  count(Location, run_prefix..exp.) %>%
  mutate_at(c("Location", "run_prefix..exp."), function(x) replace(x, x=='', "unknown")) %>%
  pivot_wider(names_from = run_prefix..exp., values_from= n)
```

Does run_prefix have any relationship with SAMPLE_TYPE?
```{r}
data %>% 
  count(SAMPLE_TYPE, run_prefix..exp.) %>%
  mutate_at(c("SAMPLE_TYPE", "run_prefix..exp."), function(x) replace(x, x=='', "unknown")) %>%
  pivot_wider(names_from = run_prefix..exp., values_from= n)
```

Does run_prefix have any relationship with AvgSpotLen?
```{r}
data %>% 
  count(AvgSpotLen, run_prefix..exp.) %>%
  mutate_at(c("AvgSpotLen", "run_prefix..exp."), function(x) replace(x, x=='', "unknown")) %>%
  pivot_wider(names_from = run_prefix..exp., values_from= n)
```
The samples with prefix "MiSeq" have avg spot lengths of 150-151. All of the ones that start with "HiSeq" have ave spot length <150.  I'm not sure why there is a relationship between different HiSeq prefix sets and different spot lengths.

### Date = COLLECTION_TIMESTAMP

Is there any difference between "Date" and "COLLECTOIN_TIMESTAMP"
```{r}
table(data$Date == gsub(" 12:00", "", data$COLLECTION_TIMESTAMP))
```
COLLECTOIN_TIMESTAMP includes " 12:00" at the end of the Date value.



### run_prefix vs Date

Does run_prefix have any relationship with Date?
```{r}
df5 = df2 %>% 
  mutate_at(c("study_day"), function(x) round(x, -1)) %>%
  count(study_day, run_prefix..exp.) %>%
  mutate_at(c("study_day", "run_prefix..exp."), function(x) replace(x, x=='', "unknown")) 

ggplot(df5, aes(x=as.numeric(study_day), y=run_prefix..exp., color=n, fill=n)) + 
  geom_tile()
```

```{r}
# df5 %>% pivot_wider(names_from = run_prefix..exp., values_from= n)
```

### env_biome

Is env_biome interesting?
```{r}
table(data$env_biome)
```

Nope. not interesting.

### env_feature

Is env_feature interesting?
```{r}
feats = split(data, f=data$env_feature)
lapply(feats, function(df) unique(df$SAMPLE_TYPE))
```
No, env_feature is a grouping or category of SAMPLE_TYPE; SAMPLE_TYPE is a sub-type of env_feature. not interesting.

### env_material

Is env_material interesting?
```{r}
mats = split(data, f=data$env_material)
lapply(mats, function(df) unique(df$SAMPLE_TYPE))
```
Not really, env_material is a grouping or category of SAMPLE_TYPE; SAMPLE_TYPE is a sub-type of env_material not interesting. But, a helpful category, so lets keep it.

### surcat

Is surcat interesting?
```{r}
surcats = split(data, f=data$surcat)
lapply(surcats, function(df) unique(df$SAMPLE_TYPE))
```
Yes! surcat is interesting because it distinguishes staff skin from patient skin for "Hand" and "Nose" samples. (that was implied by location...but not explicit)

Does that correspond with Location?
```{r}
lapply(surcats[c("Patient_Skin", "Staff_Skin")], function(df) unique(df$Location))
```
yes! 

## General review of columns

Is anything else interesting?
```{r}
for (col in names(data)){
  vals = unique(data[,col])
  if (length(vals)==1){
    message("all rows: ", col, " = ", vals)
  } 
  else if (length(vals) > 6 ) {
    message(col, " (n=", length(vals), ")")
    print(vals[1:6])
  }else {
    message(col, " (n=", length(vals), ")")
    print(vals)
  }
}
```

I think we grabbed all the info that we might care about. :)


```{r}
sessionInfo()
```

