---
title: "Figure_1A_etc"
output: html_document
date: "2023-06-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# goal

Here, we want to recreate Figure 1A of Lax et al 2017.

In addition to looking at the floor samples like they did, we will look at each sample type separately.  In Figure 1A, they use room floor and station floor samples.

# libs/input/output

## libraries

```{r}
library(phyloseq); packageVersion("phyloseq")
# library(Biostrings); packageVersion("Biostrings")
library(ggplot2); packageVersion("ggplot2")
```


## metadata

```{r}
metaReduced = dir("../..", pattern="metadata_PRJEB14474", full.names = T, include.dirs = T)
metaFile = file.path(metaReduced, "output", "PRJEB14474_SraRunTable_reduced.txt")
meta = read.delim2(metaFile)
row.names(meta) = meta$ID
message("Read metadata from: ", metaFile)
```

We will need to describe days as being "pre-opening" or "post-opening".  This information is held is "study_day". Make a new variable that is a simple category instead of a numeric.
```{r}
meta$opening = "before opening"
meta$opening[meta$study_day > 0] = "after opening"
```


## counts data

```{r}
DADA2_part2 = dir("../..", pattern="DADA2_part2", full.names = T, include.dirs = T)
countsFile = file.path(DADA2_part2, "output", "asv-counts.txt")
counts = read.delim2(countsFile, row.names = 1)
dim(counts)

counts = counts[rowSums(counts) > 0,]
dim(counts)
```

## taxonomy

```{r}
assignTax = dir("../..", pattern="assign_ASV_taxonomy", full.names = T, include.dirs = T)
taxaFile = file.path(assignTax, "output/silva/asvTaxaTable_silva-v138.txt")
taxa = read.delim2(taxaFile, row.names = "asvID")
# drop the ASV column
taxaLim = taxa[,grep("ASV", names(taxa), invert = T, value = T)]
taxaMat = as.matrix(taxaLim)
```


## direct output

```{r}
# tmpDir = "../temp"
# suppressWarnings(dir.create(tmpDir, recursive = T))
# 
outDir = "../output"
suppressWarnings(dir.create(outDir, recursive = T))
```

# Ordination function

In the legend, Figure 1A is described as:

> (A) Principal coordinate analysis (PCoA) of all floor samples based on weighted UniFrac distance and colored by whether they were taken before or after the hospital’s opening.

```{r}
# alt syntax: 
# doPCoA = function(counts=counts, meta=meta, taxaMat=taxaMat, sampleType="Floor"){

doPCoA = function(sampleType="Floor"){
  sampleSet = meta[meta$SAMPLE_TYPE %in% sampleType, "ID"]
  countslim = counts[sampleSet,]
  
  ps <- phyloseq(otu_table(counts[sampleSet,], taxa_are_rows=FALSE), 
                 sample_data(meta[sampleSet,]), 
                 tax_table(taxaMat))
  ps.prop <- transform_sample_counts(ps, function(otu) otu/sum(otu))
  ord <- ordinate(ps.prop, method="PCoA", distance="bray") # TODO: switch to unifrac to match paper
  po = plot_ordination(ps.prop, ord, shape="surcat", color="opening", title=paste0("bray PCoA - ", sampleType))
  return(po)
}
floor = doPCoA(sampleType="Floor")
floor
ggsave(file.path(outDir, "floor_bray_PCoA.png"))
```

Do the same for each sample type.

```{r}
sampleTypes = unique(meta$SAMPLE_TYPE)
plotList = as.list(sampleTypes)
names(plotList) = sampleTypes

for (type in sampleTypes){
  aplot = doPCoA(sampleType=type)
  plotList[[type]] = aplot
  show(aplot)
}
```

Save images to a pdf.
```{r}
pdf(file.path(outDir, "allTypes_unifrac_PCoA.pdf"))

for (type in sampleTypes){
  show(plotList[[type]])
}

dev.off()
```












```{r}
sessionInfo()
```

