---
title: "pick-pairs"
output: html_document
date: "2023-11-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Make pairs sets of samples.  We are aiming to make one or more sets of samples. Withen each set, we should have pairs of samples that are definitively more like each other than the others.

Libraries
```{r}
tellme <- function(name){print(paste0("Package ", name, " version: ", packageVersion(name)))}

suppressPackageStartupMessages(library(dplyr))
tellme("dplyr")
```

# Importing data

Read our reduced meta data.
```{r}
metaDir = dir("../../", pattern="review_metadata_PRJEB14474", full.names = T)
meta = read.delim(file.path(metaDir, "output", "PRJEB14474_SraRunTable_reduced.txt"))
dim(meta)
```



Read the patientID mapping file.
```{r}
pm = read.delim("../../data/NOT_IN_GIT/meta/PatientID_Mapping.txt")
dim(pm)
```

Join the sra metadata to the patientID mapping table.
```{r}
df = merge(pm, meta, by.x="SampleID", by.y="internalID")
dim(df)
```

Remove trip hazards.
```{r}
# remove any technical reps.
df = df %>% 
  arrange(desc(Bases)) %>%
  filter(!duplicated(host_subject_id)) 

# make sure this is a numeric column
df$Days_Post_Hospital_Opening = as.numeric(df$Days_Post_Hospital_Opening)

# "null" is not a PatientID
df$PatientID[df$PatientID=="null"] = NA
```

## pairs

patient nose
```{r}
getPairs = function(dfs, idColumn="PatientID", maxDaysApart = NA){
  # dfs - data frame with samples as rows
  dfs$pairID = dfs[,idColumn]
  pairset = data.frame(PairID="starter", 
                       ID.Left=NA, 
                       ID.Right=NA,
                       daysApart=NA)
  for (pid in unique(dfs$pairID)){
    dfmini = dfs %>% filter(pairID==pid)
    if (nrow(dfmini) > 1){
      dfmini = dfmini %>% arrange(Days_Post_Hospital_Opening)
      firstDay = dfmini[1, "Days_Post_Hospital_Opening"]
      secondDay = dfmini[2, "Days_Post_Hospital_Opening"]
      daysApart = secondDay - firstDay
      pairset = rbind(pairset, 
                      c(pid, 
                        dfmini[1, "ID"], 
                        dfmini[2, "ID"],
                        daysApart=daysApart))
    }
  }
  # remove empty starter row
  pairset = pairset %>% filter(PairID != "starter") %>% select(-PairID)
  # limit days separation
  if (!is.na(maxDaysApart)){
    pairset = pairset %>% 
      filter(daysApart <= maxDaysApart) %>%
      select(-daysApart)
  }
  pairset
}

pairs = df %>% 
  filter(Surface == "Patient Nose") %>%
  getPairs()

dim(pairs)
head(pairs)
```

_Seeing 19 rows_

Most of those (16/19) are only one day apart.  It will probably be the same for other patient samples, so lets only use the ones that are one day apart. Save to a file.

Save all files to a "pairs" folder.
```{r}
dir.create("../pairs")
my.write.table <- function(x, file, ...){
  write.table(x=x, 
              file=file.path("../pairs", file), 
              sep="\t", quote=F, row.names = F, ...)
  return(dim(x))
}
```


```{r}
df %>% 
  filter(Surface == "Patient Nose") %>%
  getPairs(maxDaysApart=1) %>%
  my.write.table(file="pairs_Patient-Nose.txt")

df %>% 
  filter(Surface == "Patient Axilla") %>%
  getPairs(maxDaysApart=1) %>%
  my.write.table(file="pairs_Patient-Axilla.txt")

df %>% 
  filter(Surface == "Patient Hand") %>%
  getPairs(maxDaysApart=1) %>%
  my.write.table(file="pairs_Patient-Hand.txt")

df %>% 
  filter(Surface == "Staff Hand") %>%
  getPairs(maxDaysApart=30, idColumn="Room_Number") %>%
  my.write.table(file="pairs_Staff-Hand.txt")

df %>% 
  filter(Surface == "Staff Nose") %>%
  getPairs(maxDaysApart=60, idColumn="Room_Number") %>%
  my.write.table(file="pairs_Staff-Nose.txt")
```









```{r}
sessionInfo()
```

